A Formalization of Typed Lua

### 2015-08-18 doc/papers/dls15/paper.tex
change point size, squeeze figures

\documentclass[preprint]{sigplanconf}

@@

\begin{figure}[t]
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}
\caption{Building an object piece-by-piece}
\label{fig:newpoint}
\end{figure}

@@

\begin{figure}[t]
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}
\caption{Overloading on multiple returns}
\label{fig:over}
\end{figure}

@@

\begin{figure*}[t]
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{F:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; F & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
\caption{Special Types}
\label{fig:special}
\end{figure*}

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Pair}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash P_{1} \subtype P_{2}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\begin{array}{c}
\mylabel{S-Variadic}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
{\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-VarExpand}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
	\senv \vdash F_{1}{*} \subtype P_{2}}
{\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VarDrop}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
	\senv \vdash P_{1} \subtype F_{2}{*}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]
\caption{Subtyping for second-level types}	
\label{fig:subtuples}
\end{figure}

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;\
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{fixed|closed} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{unique} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{open} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{unique} \subtype
	\{\overline{F^\prime:V^\prime}\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{open} \subtype
	\{\overline{F^\prime:V^\prime}\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{fixed} \subtype
	\{\overline{F^\prime:V^\prime}\}_{fixed}}
\end{array}
\end{array}
\]
\caption{Subtyping for table types}
\label{fig:tablesub}
\end{figure}

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash F_{2} \subtype F_{1}}
{\senv \vdash F_{1} \subtype_{c} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{u} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{u} F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype F}
{\senv \vdash \Nil \subtype_{o} F}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype F}
{\senv \vdash \Nil \subtype_{o} \Const \; F}
\end{array}
\end{array}
\]
\caption{Subtyping for fields}
\label{fig:subfield}
\end{figure}

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash F_{1} \subtype F_{2}}
{\senv \vdash \mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash F_{1} \subtype [x \mapsto \mu x.F_{2}]F_{2}}
{\senv \vdash F_{1} \subtype \mu x.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.F_{1}]F_{1} \subtype F_{2}}
{\senv \vdash \mu x.F_{1} \subtype F_{2}}
\end{array}
\end{array}
\]
\caption{Subtyping for recursive types}
\label{fig:subrec}
\end{figure}

\begin{figure*}[t]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = e \; \mathbf{in} \; s & \textit{recursive declaration} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
	& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
Lua. The first way is by assignment to a non-existing
field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
$x$ and $y$ is $\Number$, we can translate the first three lines
of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;$}\\
& \multicolumn{1}{l}{$point[``y"] \; {<}\Number{>} = y$}
\end{tabular}
\end{center}

@@

We could add both fields with a single assignment statement,
as in the following example:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
		& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>},point[``y"] \; {<}\Number{>} = x,y;$}
	\end{tabular}
\end{center}

In this case, the first coercion expression would still change the type of $point$ to $\{``x":\Number\}_{unique}$, but the
second would change it to $\{``y":\Number\}_{unique}$.
Merging the two yields $\{``x":\Number, ``y": \Number\}_{unique}$
through the third case of function $merget$ defined in Figure~\ref{fig:funcs}.

@@

{\small

>>>>

\documentclass[10pt]{sigplanconf}

@@

\begin{figure}[t]
{\footnotesize
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}
}
\caption{Building an object piece-by-piece}
\label{fig:newpoint}
\end{figure}

@@

\begin{figure}[t]
{\footnotesize
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}
}
\caption{Overloading on multiple returns}
\label{fig:over}
\end{figure}

@@

\begin{figure}[t]
{\scriptsize
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{F:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
&  \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \Boolean \ |\  \Integer \ |\ \Number\ |\ \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
}
\caption{Type Language}
\label{fig:typelang}
\end{figure}

\begin{figure}[t]
{\footnotesize
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; F & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
}
\caption{Special Types}
\label{fig:special}
\end{figure}

@@

\begin{figure}[t]
{\footnotesize
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Pair}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash P_{1} \subtype P_{2}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\begin{array}{c}
\mylabel{S-Variadic}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
{\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-VarExpand}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
	\senv \vdash F_{1}{*} \subtype P_{2}}
{\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VarDrop}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
	\senv \vdash P_{1} \subtype F_{2}{*}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]
}
\caption{Subtyping for second-level types}	
\label{fig:subtuples}
\end{figure}

@@

\begin{figure}[t]
{\footnotesize
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;\
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{fixed|closed} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{unique} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{open} \subtype
	\{\overline{F^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{unique} \subtype
	\{\overline{F^\prime:V^\prime}\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{F:V}\}_{open} \subtype
	\{\overline{F^\prime:V^\prime}\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\forall i \; \exists j \;\;\;
	\senv \vdash F_{i} \subtype F_{j}' \;
	\senv \vdash F_{j}' \subtype F_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\end{array}}
{\senv \vdash \{\overline{F:V}\}_{fixed} \subtype
	\{\overline{F^\prime:V^\prime}\}_{fixed}}
\end{array}
\end{array}
\]
}
\caption{Subtyping for table types}
\label{fig:tablesub}
\end{figure}

\begin{figure}[t]
{\footnotesize
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash F_{2} \subtype F_{1}}
{\senv \vdash F_{1} \subtype_{c} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{u} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{u} F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype F}
{\senv \vdash \Nil \subtype_{o} F}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype F}
{\senv \vdash \Nil \subtype_{o} \Const \; F}
\end{array}
\end{array}
\]
}
\caption{Subtyping for fields}
\label{fig:subfield}
\end{figure}

@@

\begin{figure}[t]
{\footnotesize
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash F_{1} \subtype F_{2}}
{\senv \vdash \mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash F_{1} \subtype [x \mapsto \mu x.F_{2}]F_{2}}
{\senv \vdash F_{1} \subtype \mu x.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.F_{1}]F_{1} \subtype F_{2}}
{\senv \vdash \mu x.F_{1} \subtype F_{2}}
\end{array}
\end{array}
\]
}
\caption{Subtyping for recursive types}
\label{fig:subrec}
\end{figure}

\begin{figure}[t]
{\scriptsize
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = e \; \mathbf{in} \; s & \textit{recursive declaration} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{function call}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{one result}\\
	l ::= & & \textsc{lvalues:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
	& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literals:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{calls:}\\
	& \;\; e(el) & \textit{function call}\\
	& | \; e{:}n(el) & \textit{method call}\\
	f ::= & & \textsc{functions:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
}
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure}

@@

There are two ways the type of a {\em unique} and {\em open} tables can change in Typed
Lua. The first way is by assignment to a non-existing
field. Let us go back to the example shown in Figure~\ref{fig:newpoint}. Assuming that the type of the
$x$ and $y$ is $\Number$, we can translate the first three lines
of function {\tt new\_point} to the abstract syntax of Figure~\ref{fig:syntax}:
{\small
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>} = x;$}\\
& \multicolumn{1}{l}{$point[``y"] \; {<}\Number{>} = y$}
\end{tabular}
\end{center}
}

@@

We could add both fields with a single assignment statement,
as in the following example:
{\small
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \; point: \{\}_{unique} = \{\} \; \mathbf{in}$}\\
		& \multicolumn{1}{l}{$point[``x"] \; {<}\Number{>},point[``y"] \; {<}\Number{>} = x,y;$}
	\end{tabular}
\end{center}
}
In this case, the first coercion expression would still change the type of $point$ to $\{``x":\Number\}_{unique}$, but the
second would change it to $\{``y":\Number\}_{unique}$.
Merging the two yields $\{``x":\Number, ``y": \Number\}_{unique}$
through the third case of function $merget$ defined in Figure~\ref{fig:funcs}.

@@

{\footnotesize
